---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 3 - Due date 02/03/26"
author: "Sylvia Hipp"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp25.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

## Questions

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2025 Monthly Energy Review. This time you will work only with the following columns: 
**Total Renewable Energy Production**; and 
**Hydroelectric Power Consumption**.

Create a data frame structure with these two time series only.

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r message=FALSE}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(Kendall)

# additional packages
library(readxl)
library(lubridate)
library(ggplot2)
library(glue)
library(tidyverse)
```

Read in the raw data
```{r}
filepath <- "./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx"

# load the data
energy_data_raw <- read_excel(filepath, skip = 11)

# read in column names
colnames <- read_excel(filepath, skip = 10) %>% names()
names(energy_data_raw) <- colnames  # assign colnames to raw dataset
```

Basic data processing 
```{r}
# clean up names and select columns of interest
energy_data <- energy_data_raw %>% 
  janitor::clean_names() %>%   
  select(month, total_renewable_energy_production, hydroelectric_power_consumption)
```


##Trend Component

### Q1

For each series (Total Renewable Production and Hydroelectric Consumption) create three plots arranged in a row (side-by-side): (1) time series plot, (2) ACF, (3) PACF. Use cowplot::plot_grid() to place them in a grid.

```{r}
# function to create all three plots for a given series
create_plots <- function(series, description){
  time_series_plot <- ggplot(energy_data, 
                             aes(x = month, y = series)) + 
    geom_line() + 
    scale_x_date(expand = c(0,0)) +   # compress x axis
    theme_classic() + 
    labs(x = "Month", 
         y = glue("Total {description} (Trillion Btu)"), 
         title = glue("Time Series of\n{description}")) 
  
  acf_plot <- ggAcf(series, lag.max = 40) +
    theme_classic() +
    labs(title = glue("{description} ACF"))
  
  pacf_plot <- ggPacf(series, lag.max = 40) +
    theme_classic() +
    labs(title = glue("{description} PACF"))
  
  cowplot::plot_grid(time_series_plot, acf_plot, pacf_plot, 
                     ncol = 3)
}

# Series 1: Renewable Production
create_plots(energy_data$total_renewable_energy_production, 
             "Renewable\nProduction")

# Series 2: Hydroelectric Consumption
create_plots(energy_data$hydroelectric_power_consumption, 
             "Hydroelectric\nConsumption")
```


### Q2

From the plot in Q1, do the series Total Renewable Energy Production and Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

> Answer: Renewable Energy Production appears to have a strong positive linear trend over time, although this trend seems to stall over the period 1985–200. Hydroelectric Power Consumption does not not appear to have a strong trend. There may be a slight negative trend, but this is difficult to tell apart from the seasonal variation. 

### Q3

Use the *lm()* function to fit a linear trend to the two time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}
t <- c(1:nrow(energy_data))

# Total Renewable Energy Production
linear_model_renewable <- lm(energy_data$total_renewable_energy_production~t)
summary(linear_model_renewable)

# save coefficients
renewable_b0 <- linear_model_renewable$coefficients[1] %>% as.numeric()
renewable_b1 <- linear_model_renewable$coefficients[2] %>% as.numeric()

# Hydroelectric Power Consumption
linear_model_hydro <- lm(energy_data$hydroelectric_power_consumption~t)
summary(linear_model_hydro)

# save coefficients
hydro_b0 <- linear_model_hydro$coefficients[1] %>% as.numeric()
hydro_b1 <- linear_model_hydro$coefficients[2] %>% as.numeric()

```

> Interpretation: The linear trend fit to the renewable energy production data is positive with an adjusted r-squared value of 0.8201, meaning 82% of variation in this series can be explained by the linear trend over time. These results are significant with a p-value less than 0.05. The linear trend fit to the hydroelectric consumption data is negative with an adjusted r-squared value of 0.02343, meaning only 2.34% of variation in the data can be explained by this trend. This trend is significant with a p-value < 0.05. 

### Q4

Use the regression coefficients to detrend each series (subtract fitted linear trend). Plot detrended series and compare with the original time series from Q1. Describe what changed.

```{r}
# function to detrend series
detrend_series <- function(series, b0, b1, description){
  # subtract linear trend
  series_detrend <- series - (b0 + b1*t)
  
  # x-axis values
  months <- seq(first(energy_data$month), last(energy_data$month),   
                by = "month")  
  
  # combine values into a df to plot
  df_to_plot <- data.frame("month" = months, 
                           "series" = series, 
                           "series_detrend" = series_detrend) %>% 
    pivot_longer(cols = c("series", "series_detrend"), 
                 names_to = "series") %>% 
    mutate(series = factor(series, 
                           levels = c("series", "series_detrend"), 
                           labels = c(glue("Observed {description}"), 
                                      glue("Detrended {description}")))) 
  # plot both time series
  time_series_plot <- ggplot(df_to_plot) + 
    geom_line(aes(x = month, y = value, color = series)) + 
    theme_classic() + 
    theme(legend.position = "bottom") + 
    labs(x = "Month", 
         y = glue("{description} (Trillion Btu)"), 
         title = glue("Detrended {description}"), 
         color = NULL)
    
  return(list(series_detrend, time_series_plot))
}

# detrend renewable energy production
renewable_detrended <- detrend_series(energy_data$total_renewable_energy_production, 
                                      b0 = renewable_b0, b1 = renewable_b1, 
                                      description = "Renewable Energy Production")

renewable_detrend_series <- renewable_detrended[[1]] # save detrended series object
renewable_detrended[[2]]   # show plot

# detrend hydroelectric power consumption
hydro_detrended <- detrend_series(energy_data$hydroelectric_power_consumption, 
                                  b0 = hydro_b0, b1 = hydro_b1, 
                                  description = "Hydroelectric Power Consumption")

hydro_detrend_series <- hydro_detrended[[1]]   # save detrended series object
hydro_detrended[[2]]   # show plot
```

> Answer: Both detrended series no longer have a trend component and their values are centered around 0. For hydroelectric power consumption, the shape of the detrended series appears the same and only the values have been shifted down. The series of renewable energy production appears to have been flattened out. 

### Q5

Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side to make it easier to compare. Did the plots change? How?

```{r}
plot_detrend_acf <- function(series, detrend_series, description){
  plot_acf <- ggAcf(series) + 
    theme_classic() + 
    labs(title = glue("ACF of {description}"))
  
  plot_pacf <- ggPacf(series) + 
    theme_classic() + 
    labs(title = glue("PACF of {description}"))
  
  plot_acf_detrend <- ggAcf(detrend_series) + 
    theme_classic() + 
    labs(title = glue("ACF of Detrended {description}"))
  
  plot_pacf_detrend <- ggPacf(detrend_series) + 
    theme_classic() + 
    labs(title = glue("PACF of Detrended {description}"))
  
  cowplot::plot_grid(plot_acf, plot_pacf, plot_acf_detrend, plot_pacf_detrend, 
            ncol = 2, nrow = 2)
}
# Renewable Energy Production
plot_detrend_acf(energy_data$total_renewable_energy_production, 
                 renewable_detrend_series, 
                 "Renewable Energy Production")

# Hydroelectric Power Consumption
plot_detrend_acf(energy_data$hydroelectric_power_consumption, 
                 hydro_detrend_series, 
                 "Hydroelectric Power Consumption")
```

> Answer: There is a slight change in the ACF and PACF for the renewable energy production detrended series. For these data, the seasonal component is a bit more pronounced in the ACF and PACF of the detrended series where there appears to be a slight spike in autocorrelation at around 12 lags. There appears to be no change in the ACF and PACF for the detrended hydroelectric power consumption series. 

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6

Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

> Answer: For renewable energy production, there is no apparent seasonal component to the time series and ACF plots. For hydroelectric power consumption, there is a very clear seasonal component to both the ACF and time series plots. 

### Q7

Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results, which series have a seasonal trend? Do the results match you answer to Q6?

```{r}
# create time series objects
start <- c(first(year(energy_data$month)), first(month(energy_data$month)))

renewable_ts <- ts(energy_data$total_renewable_energy_production, 
                   start = start, frequency = 12)
hydro_ts <- ts(energy_data$hydroelectric_power_consumption, 
               start = start, frequency = 12)

# create seasonal dummies
dummies <- seasonaldummy(renewable_ts)

# fit seasonal means model to both series
# Renewable Energy Production
renewable_seasonal_model <- lm(renewable_ts ~ dummies)
summary(renewable_seasonal_model)

# Hydro Power Consumptions
hydro_seasonal_model <- lm(hydro_ts ~ dummies)
summary(hydro_seasonal_model)
```

> Answer

### Q8

Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q9

Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. Did the plots change? How?
