---
title: "ENV 797 - Time Series Analysis for Energy and Environment Applications | Spring 2026"
subtitle: "Assignment 6 - Due date 02/27/26"
author: "Sylvia Hipp"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp26.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

R packages needed for this assignment: "ggplot2", "forecast", "tseries" and "sarima". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(cowplot)
library(sarima)
```

This assignment has general questions about ARIMA Models.

## Q1

Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models:

* AR(2)

> Answer: The ACF plot of an AR(2) model will show exponential decay over time. The PACF plot will show rapid decay and have insignificant values after 2 lags, which is indicative of a 2nd order AR model.

* MA(1)

> Answer: The ACF plot of an MA(2) model will show rapid decay and have insignificant values after 1 lag, which is indicative of the 1st order MA model. The PACF plot will show exponential decay over time. 

## Q2

Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$.

(a) Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use the `arima.sim()` function in R to generate $n=100$ observations from each of these three models. Then, using `autoplot()` plot the generated series in three separate graphs.

```{r}
arma_10 <- arima.sim(n=100, model=list(order=c(1,0,0), ar=0.6))
arma_01 <- arima.sim(n=100, model=list(order=c(0,0,1), ma=0.9))
arma_11 <- arima.sim(n=100, model=list(order=c(1,0,1), ar=0.6, ma=0.9))

autoplot(arma_10) + labs(title = "ARMA(1,0)")
autoplot(arma_01) + labs(title = "ARMA(0,1)")
autoplot(arma_11) + labs(title = "ARMA(1,1)")
```

(b) Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use `cowplot::plot_grid()`).


```{r}
plot_grid(
  ggAcf(arma_10) + labs(title="ARMA(1,0)"), 
  ggAcf(arma_01) + labs(title="ARMA(0,1)"), 
  ggAcf(arma_11) + labs(title="ARMA(1,1)"), 
  nrow = 3
)
```

(c) Plot the sample PACF for each of these models in one window to facilitate comparison.

```{r}
plot_grid(
  ggPacf(arma_10) + labs(title="ARMA(1,0)"), 
  ggPacf(arma_01) + labs(title="ARMA(0,1)"), 
  ggPacf(arma_11) + labs(title="ARMA(1,1)"), 
  nrow = 3
)
```

(d) Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be able identify them correctly? Explain your answer.

** ASK IN OFFICE HOURS**

> Answer: I believe I would be able to identify each model. For the ARMA(1,0) there is exponential decay of the ACF and a sharp drop off after the first lag of the PACF. The opposit is true for ARMA(0,1) -- there is a sharp drop off after one lag of the ACF and exponential decay of the PACF. The ARMA(1,1) would be a bit more difficult. Both the ACF and PACF become insignficicant after about two lags, but there are a few lags that are very marginally significant. While I recognize this is rapid decay because I know its an ARMA model, I may have more trouble recognizing it without that information. You would not be able to determine the order of this model just looking at the ACF/PACF. 

(e) Compare the PACF values R computed with the values you provided for the lag 1 correlation coefficient, i.e., does $\phi=0.6$ match what you see on PACF for ARMA(1,0), and ARMA(1,1)? Should they match?

> Answer: Yes, the value of the PACF of the first lag for both models appears to be around 0.6.

(f) Increase number of observations to $n=1000$ and repeat parts (b)-(e).

```{r}
arma_10 <- arima.sim(n=1000, model=list(order=c(1,0,0), ar=0.6))
arma_01 <- arima.sim(n=1000, model=list(order=c(0,0,1), ma=0.9))
arma_11 <- arima.sim(n=1000, model=list(order=c(1,0,1), ar=0.6, ma=0.9))

autoplot(arma_10) + labs(title = "ARMA(1,0)")
autoplot(arma_01) + labs(title = "ARMA(0,1)")
autoplot(arma_11) + labs(title = "ARMA(1,1)")

plot_grid(
  ggAcf(arma_10) + labs(title="ARMA(1,0)"), 
  ggAcf(arma_01) + labs(title="ARMA(0,1)"), 
  ggAcf(arma_11) + labs(title="ARMA(1,1)"), 
  nrow = 3
)

plot_grid(
  ggPacf(arma_10) + labs(title="ARMA(1,0)"), 
  ggPacf(arma_01) + labs(title="ARMA(0,1)"), 
  ggPacf(arma_11) + labs(title="ARMA(1,1)"), 
  nrow = 3
)
```

** ASK IN OFFICE HOURS**
> Answer: Ask what ACF / PACF should look like for ARMA, my understanding was that it was sharp drop off for both PACF and ACF based on the slides / class, but that's not what I'm seeing here

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

(a) Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

** ASK IN OFFICE HOURS** --> are we able to tell anything about the differencing? Also isn't there supposed to be a constant term?
> Answer: ARIMA(1, 0, 1)(1, 0, 0). I assume that this model is on a stationary time series and that d=0, though it's not clear from the equation.

(b) Also from the equation what are the values of the parameters, i.e., model coefficients.

** CONFIRM PHI IS NEGATIVE FOR SAR**
> Answer: phi for the first order AR is 0.7, theta for the first order MA is 0.1, and phi for the first order SAR is -0.25.

## Q4

Simulate a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using the `sim_sarima()` function from package `sarima`. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot the generated
series using `autoplot()`. Does it look seasonal?

```{r}
# why is this not automatically a time series? 
sarima_0110 <- sim_sarima(n=1000, 
                          model=list(nseasons=12, ma=0.5, sar=0.8)) %>% 
  ts(frequency = 12)  # ASK IN OFFICE HOURS

autoplot(sarima_0110) + 
  scale_x_continuous(breaks=seq(0, 1000, by = 12))
```

## Q5

Plot ACF and PACF of the simulated series in Q4. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
Acf(sarima_0110)
Pacf(sarima_0110)

```
